Reintroduction of resistant frogs facilitates landscape-scale recovery in the presence of a lethal fungal disease

### 2024-01-06 doc/manuscript/translocation_pnas.tex
Minor edits to manuscript, figures, and latex files

\affil[a]{Sierra Nevada Aquatic Research Laboratory, University of California, Mammoth Lakes, CA, 93546}
\affil[b]{Earth Research Institute, University of California, Santa Barbara, CA, 93106-3060}
\affil[c]{School of Natural Resources, University of Tennessee Institute of Agriculture, Knoxville, TN, 37996}
\affil[d]{Department of Environmental Science, Policy, and Management, University of California - Berkeley, Berkeley, CA, 94720-3114}
\affil[e]{Museum of Vertebrate Zoology, University of California - Berkeley, Berkeley, CA, 94720-3160}
\affil[f]{Earth Lab, University of Colorado, Boulder, CO, 80303}\
\affil[g]{Current affiliation: Planet, San Francisco, CA, 94107}
\affil[h]{Current affiliation: Ginkgo Bioworks, Boston, MA, 02210}
\affil[i]{Resources Management and Science, Yosemite National Park, El Portal, CA, 95318}

@@

\dropcap{H}uman activities are increasingly impacting global biodiversity
\citep{ceballos2015}, with important implications for ecosystem
resilience and human welfare \citep{naeem2009}. One consequence of human
alteration of the biosphere is an increase in emerging infectious
diseases \citep{jones2008, fisher2012}. Such diseases pose a severe
threat to wildlife populations \citep{daszak2000}, and have caused
dramatic declines and extinctions in a wide range of taxa, including
echinoderms, mammals, birds, and amphibians
\citep{hewson2014, samuel2015, scheele2019, cunningham2021}. Amphibians
are experiencing particularly devastating impacts of disease due to the
recent emergence and global spread of the highly virulent amphibian
chytrid fungus, \emph{Batrachochytrium dendrobatidis} (Bd)
\citep{luedtke2023, scheele2019}. By one estimate, hundreds of species
have experienced Bd-caused declines, and numerous susceptible taxa are
extinct in the wild \citep{scheele2019}. These impacts to global
biodiversity may be unprecedented, and highlight the importance of
understanding mechanisms of species persistence in the presence of
emerging diseases \citep{russell2020}.

Evidence of natural recovery in the many Bd-impacted amphibian
populations is surprisingly rare \citep[for notable exceptions,
see][]{scheele2017, voyles2018, knapp2016}, suggesting that
disease-caused declines will be difficult to reverse. This apparent low
resilience to disease effects may be due to the limited ability of many
amphibians to develop Bd resistance and/or tolerance, which in turn,
could also lessen the effectiveness of potential Bd mitigation
strategies. Following pathogen arrival in a host population, resistance
(ability to limit pathogen burden) and tolerance (ability to limit the
harm caused by a particular burden) are key mechanisms to reduce disease
impacts \citep{raberg2009} and facilitate population persistence and
recovery \citep{brannelly2021}. Host immunity and evolution both play
important roles in the development of resistance and tolerance, and
utilizing these factors would seem a promising approach to developing
effective strategies to mitigate disease impacts in the wild
\citep{garner2016, woodhams2011}. However, several aspects of the
amphibian-Bd system present difficult obstacles, including (i) the
general inability of amphibians to mount an effective immune response
against Bd infection \citep{rosenblum2012, fites2013, grogan2018a}, and
(ii) the apparent rarity of evolution of more resistant/tolerant
genotypes \citep[but see][]{savage2016, grogan2018b}. These factors
suggest that reintroduction of amphibians into sites to reestablish
populations extirpated by Bd will often result not in population
recovery, but instead in the rapid reinfection and mortality of the
introduced animals and/or their progeny
\citep{hammond2021, knapp2022, stockwell2008, scheele2021}. If true, the
future of many amphibian species threatened by Bd appears bleak.

The mountain yellow-legged (MYL) frog, composed of the sister species
\emph{Rana muscosa} and \emph{Rana sierrae} \citep{vredenburg2007}, is
emblematic of the global declines of amphibians caused by Bd
\citep{scheele2019}. Once the most common amphibian in the high
elevation portion of California's Sierra Nevada mountains
\citep[USA,][]{grinnell1924}, during the past century this frog has
disappeared from more than 90\% of its historical range
\citep{vredenburg2007}. Due to the severity of its decline and the
increasing probability of extinction, both species are now listed as
``endangered'' under the U.S. Endangered Species Act. In the Sierra
Nevada, this decline was initiated by the introduction of non-native
trout into the extensive historically-fishless region
\citep{bradford1989, knapp2000} starting in the late 1800s. The arrival
of Bd in the mid-1900s and its subsequent spread \citep{vredenburg2019}
caused additional large-scale population extirpations
\citep{vredenburg2010, rachowicz2006}. These Bd-caused declines are
fundamentally different from the fish-caused declines because fish
eradication is feasible \citep{knapp1998} and results in the rapid
recovery of frog populations \citep{knapp2007, vredenburg2004}. In
contrast, Bd appears to persist in habitats even in the absence of
amphibian hosts \citep{walker2007}, and therefore represents a long-term
alteration of invaded ecosystems that amphibians will need to overcome
to reestablish populations.

Despite the catastrophic impact of Bd on MYL frogs, wherein most
Bd-naive populations are extirpated following Bd arrival
\citep{vredenburg2010}, some populations have persisted after epizootics
\citep[during which Bd infection intensity on frogs is very
high,][]{briggs2010} and are now recovering
(Figure~\ref{fig-recovery-model}) \citep{knapp2016}. Frogs in these
recovering populations show reduced susceptibility to Bd infection
\citep{knapp2016}, with infection intensity (``load'') on adults
consistently in the low-to-moderate range
\citep{briggs2010, knapp2011, joseph2018}. This reduced susceptibility
is evident even under controlled laboratory conditions
\citep{knapp2016}, indicative of host resistance against Bd infection
(and not simply an effect of factors external to individual frogs, e.g.,
environmental conditions). In addition to frogs from recovering
populations having higher resistance to Bd infection than those from
naive populations, they could also have higher tolerance, but no data
are currently available to evaluate this possibility. Therefore, we
focus on resistance throughout this paper. The observed resistance of
MYL frogs could be the result of several non-mutually exclusive
mechanisms, including natural selection for more resistant genotypes
\citep{savage2016, grogan2018b}, acquired immunity \citep{grogan2018a},
and/or inherent between-population differences that pre-date Bd
exposure. The possible evolution of MYL frog resistance and subsequent
population recovery is consistent with that expected under
``evolutionary rescue'', whereby rapid evolutionary change increases the
frequency of adaptive alleles and restores positive population growth
\citep{carlson2014, searle2020}. This intriguing possibility also
suggests an opportunity to expand recovery beyond the spatial scale
possible under natural recovery by utilizing resistant frogs from
recovering populations in reintroductions to vacant habitats
(Figure~\ref{fig-recovery-model}) \citep{joseph2018, mendelson2019}.

@@

Bd loads were fairly consistent before versus after translocation, and
loads were nearly always well below the level indicative of severe
chytridiomycosis (i.e., the disease caused by Bd) and associated frog
mortality (Figure S2)
\citep{joseph2018, vredenburg2010}. Although it is possible that the
observed relatively small changes in load are a consequence of
individuals with high Bd loads dying and therefore being unavailable for
sampling during the post-translocation period, the fact that there was
little difference in pre- versus post-translocation Bd loads even in
those populations that had very high frog survival (70556, 74976 - see
below; Figure S2) suggests a true lack of
substantial change in Bd load.

The ultimate measure of reintroduction success is the establishment of a
self-sustaining population. Given that it can take years or even decades
to determine the self-sustainability of a reintroduced population
\citep[for an example in MYL frogs, see][]{joseph2018}, the use of
proxies is essential for providing shorter-term insights into
reintroduction success and the factors driving it. Results from our CMR
surveys allowed us to accurately estimate frog survival, including over
the entire CMR time series for each site and during only the 1-year
period immediately following translocation. These estimates were made
using site-specific models analyzed using the mrmr package. We use these
estimates to describe general patterns of frog survival in all
translocated cohorts, and in an among-site meta-analysis of frog
survival to identify important predictors of 1-year frog survival (e.g.,
Bd load).

@@

Results from the frog translocation study indicated that most
populations showed evidence of successful reproduction and recruitment,
and that adult survival was often relatively high (described above).
Although suggestive of population establishment, a decade or more of
surveys may be necessary to confirm that populations are in fact
self-sustaining \citep{joseph2018}. To extend our inferences of
population establishment beyond those possible from the site-specific
CMR data, we developed a population viability model. Specifically, to
test whether the observed yearly adult survival probabilities in
translocated populations were sufficient for long-term viability, we
built a stage-structured matrix model that captured known frog
demography and included demographic and environmental stochasticity. We
parameterized the model using CMR data from translocated populations and
known life history values in this system (Table S1).

@@

To determine whether MYL frog populations show genomic patterns
consistent with an evolutionary response to Bd, we compared frog exomes
(i.e., coding region of a genome) between populations with contrasting
histories of Bd exposure. Specifically, we compared frog genomes sampled
in 4 populations that have not yet experienced a Bd-caused epizootic
(``naive'') \citep{zhou2015} versus in 5 populations that experienced a
Bd epizootic during the past several decades and have since recovered to
varying degrees (``recovering''; Figure~\ref{fig-allelefrequencies})
\citep{knapp2016, vredenburg2010}. Bd-exposure histories of the 9 study
populations are based on 10-20 years of VES and Bd surveillance using
skin swabbing \citep[e.g.,][]{knapp2016, zhou2015, wilber2022}. Naive
populations are characterized by large numbers of adults (i.e, typically
1000s), Bd prevalence that is generally 0\% except during occasional Bd
failed invasions \citep[during which Bd loads remain very
low,][]{wilber2022}, and no history of Bd epizootics since we first
surveyed these populations in the late 1990s and early 2000s
\citep{zhou2015}. In contrast, recovering populations exist in an
enzootic state \citep{briggs2010}, characterized by smaller numbers of
adults (generally \textless{} 500), high Bd prevalence \citep[often
\textgreater{} 80\%,][]{knapp2011}, and, in adults, moderate Bd loads
that are typically well below the level expected to cause mortality
\citep{vredenburg2010}. Naive and recovering populations can be
identified unambiguously using these differences in Bd prevalence and
load. Finally, there is no potential for frog dispersal between the 9
study populations due to intervening distances and topography, as well
as the presence of introduced (predatory) fish and fish-induced habitat
fragmentation. (See \textbf{Supporting Information - Frog evolution in
response to Bd - Study design} for additional details regarding the
study design.)

@@

Individual frogs clustered into 3 separate groups in principal component
space (Figure S6A), and clusters reflected
the species split (i.e., \emph{R. muscosa} versus \emph{R. sierrae}) and
the strong signature of isolation-by-distance that is characteristic of
MYL frogs \citep{rothstein2020, poorten2017, byrne2023}. Importantly,
each cluster contained at least one population from both the naive and
recovering groups, allowing us to distinguish allelic associations of
individuals sampled in the 2 population types versus allelic
associations resulting from population structure and genetic drift.

@@

The splined window analysis identified 33 outlier regions for
\(\pi_{diff}\) and 58 outlier regions for \emph{F\textsubscript{ST}}
(Figure~\ref{fig-spline-manhattan}A, B). Of these, 9 regions were
outliers for both metrics (``shared regions'') and 2 of these shared
regions also contained one or more of the outlier SNPs described above.
A total of 35 annotated genes were found in the 9 shared regions. Given
this large number of genes, here we focus on those with the strongest
signal of selection and/or immune-related functions. The largest
\(\pi_{diff}\), indicative of directional selection, occurred in a 163kb
region on Contig19, 12.9Mb upstream of the RIN3 outlier SNP
(Figure~\ref{fig-spline-manhattan}C). This region contains
approximately 500 SNPs and one annotated gene called
``interferon-induced very large GTPase 1-like'' (GVINP1). Additionally,
a shared outlier region on Contig1 contained two complement factor genes
(C6 and C7). Interestingly, this region had a large negative
\(\pi_{diff}\) value, consistent with balancing selection. Finally, one
shared outlier region on Contig8 contained one outlier SNP (TCF19) and
was within 360kb of another outlier SNP (VARS)
(Figure~\ref{fig-spline-manhattan}D, Figure S7).
This region (854kb from the beginning of the outlier window to the VARS
SNP) contained a total of 8 annotated genes. In \emph{Xenopus}, five of
these genes occur in the extended major histocompatibility complex (MHC)
Class I region (FLOT1, TUBB, MDC1, CCHCR1, TCF19) and three occur in the
extended MHC Class III region (HSP70, LSM2, VARS) \citep{ohta2006}.
Therefore, this region under selection is part of the extended MHC Class
I and III complex and shows synteny with other amphibian genomes.

@@

Disease-induced population declines are decimating global biodiversity
\citep{daszak2000}, but broadly-applicable strategies to recover
affected species are generally lacking \citep[e.g.,][]{garner2016}.
Here, we tested the possibility that populations of resistant
individuals from naturally recovering populations can be used to
reestablish extirpated populations of the endangered MYL frog in the
presence of a highly virulent fungal pathogen (Bd). Our results indicate
(i) the capacity of reintroduced populations to become established and
eventually recover despite ongoing disease, (ii) that the recovering
populations are likely to persist over a 50-year period, (iii) that
there are substantial genomic differences between naive and recovering
MYL frog populations, consistent with evolutionary change in frogs
following Bd exposure, and (iv) that some of the genomic regions under
selection contain genes related to disease resistance. Collectively,
these results (Figure~\ref{fig-recovery-model}) provide a rare example
of amphibian recovery in the presence of Bd, and have important
implications for the conservation and recovery of amphibians and other
taxa worldwide that are endangered by escalating impacts from emerging
infectious diseases. In light of the generally low success rate of
amphibian reintroduction efforts \citep{dodd2005}, our success in
reestablishing MYL frog populations via translocation of resistant
individuals is striking, and even more so given that MYL frogs were
driven to near-extinction by Bd.

In the following discussion, we follow the sequence of frog recovery
described in Figure~\ref{fig-recovery-model} to structure our key
points. Previous field studies in MYL frogs show that frog-Bd dynamics
and frog survival in the presence of Bd are fundamentally different
between naive and recovering populations. Following the arrival and
establishment of Bd in previously-naive populations, adult frogs develop
high Bd loads that lead to mass die-offs \citep{vredenburg2010}. In
contrast, in recovering populations adult frogs typically have
low-to-moderate and relatively constant Bd loads and mass die-offs are
not observed \citep[see also Figure S2]{briggs2010, knapp2011}. The differences in Bd load of frogs from
naive and recovering populations are also observed in controlled
laboratory studies \citep[see Figure 4 in][]{knapp2016}, and clearly
indicate that frogs from recovering populations exhibit resistance
against Bd infection. This resistance could in theory be due to several
factors, including natural selection for more resistant genotypes,
acquired immunity, and/or inherent between-population differences that
pre-date Bd exposure, but until now evidence to evaluate the role of
evolution was lacking.

@@

Importantly, some genomic regions that we identified as under selection
are associated with cellular and immunological mechanisms known to
contribute to disease resistance, including in amphibians
\citep{zamudio2020}. For example, the MHC plays an important role in
immunity. In our study, we identified a region that shows evidence of
selection in recovering populations and contains eight genes associated
with either the MHC Class I or Class III regions. These results
corroborate numerous previous studies linking MHC genes to amphibian
resistance against Bd \citep[e.g.,][]{savage2011, bataille2015}.
Similarly, the region with the strongest indication of directional
selection (as measured by \(\pi_{diff}\)) contains the
interferon-related gene GVINP1. Several previous studies of amphibians
have found this gene to be differentially expressed during Bd infection
\citep[e.g.,][]{grogan2018b, ellison2014} and in populations differing
in Bd susceptibility \citep{grogan2018b}. This gene is also strongly
linked to disease in salmon, explaining a notable 20\% of the resistance
phenotype \citep{robledo2020, robledo2018}. We also identified a region,
characterized by high \emph{F\textsubscript{ST}} and low \(\pi_{diff}\),
that contained the complement genes C6 and C7. The complement system
plays an important role in innate immunity \citep{rieraromo2016}, and
our results could indicate that balancing selection is acting in this
region of the genome to favor a diverse set of alleles, as is known for
C6 in humans \citep{soejima2005}. Based on the analysis of individual
outlier variants, the RIN3 gene showed a consistent pattern of allele
frequency differences across all nine of the frog populations sampled in
this study, indicating consistent selection in populations distributed
across a wide geographic area. This gene is associated with immune
response and in \emph{Xenopus} is expressed during appendage
regeneration \citep{fukazawa2009}. Finally, the outlier variant with the
lowest p-value was the uncharacterized gene LOC108802036. In the genome
of another frog species, this gene is located adjacent to a type I
interferon gene (Np-IFNi2) \citep{gan2018}, and together with GVINP1
further suggests the importance of interferon-related genes in this
system. Collectively, the genes associated with these genomic
differences may confer at least some degree of resistance against Bd
infection, an attribute that may be critically important to population
reestablishment and recovery in the presence of Bd.

@@

Results from the population viability model were also encouraging. In
particular, translocated populations with \textgreater{} 50\% survival
in the first year post-translocation were predicted to have a low
probability of extinction over 50 years (probability of extinction \(<\)
0.5 when year-1 juvenile survival probability was greater than 0.10).
The viability model highlighted the important role of frog survival in
affecting long-term population viability, and allowed us to extend the
temporal scale of our study beyond the years covered by our
post-translocation surveys. These long-term forecasts are important,
given that reintroduced MYL frog populations may often take decades to
achieve our ultimate goal of self-sustainability \citep{joseph2018}.
Making well-supported projections about the long-term outcome of
reintroduction efforts from shorter-term information is critically
important to the process of adaptive management of species
reintroduction programs \citep{seddon2007}, including the one we are
carrying out for MYL frogs. Specifically, the combined results from our
reintroduction study and viability modeling indicate that survival of
frogs in the first year following translocation is an effective proxy of
longer-term survival and population viability. In addition, given the
repeatability of frog survival at a site, 1-year frog survival also
serves as an effective proxy of site quality (i.e., the ability of a
site to support high frog survival and a viable frog population over the
long term). This proxy of site quality is important in the MYL frog
system because accurately predicting the ability of a site to support a
viable frog population \emph{a priori} remains difficult, even after
conducting 24 translocations over 16 years.

>>>>

\affil[a]{Sierra Nevada Aquatic Research Laboratory, University of California, Mammoth Lakes, CA, 93546}
\affil[b]{Earth Research Institute, University of California, Santa Barbara, CA, 93106-3060}
\affil[c]{School of Natural Resources, University of Tennessee Institute of Agriculture, Knoxville, TN, 37996}
\affil[d]{Department of Environmental Science, Policy, and Management, University of California - Berkeley, Berkeley, CA, 94720-3114}
\affil[e]{Museum of Vertebrate Zoology, University of California - Berkeley, Berkeley, CA, 94720-3160}
\affil[f]{Earth Lab, University of Colorado, Boulder, CO, 80303}
\affil[g]{Current affiliation: Planet, San Francisco, CA, 94107}
\affil[h]{Current affiliation: Ginkgo Bioworks, Boston, MA, 02210}
\affil[i]{Resources Management and Science, Yosemite National Park, El Portal, CA, 95318}

@@

\dropcap{H}uman activities are increasingly impacting global biodiversity
\cite{ceballos2015}, with important implications for ecosystem
resilience and human welfare \cite{naeem2009}. One consequence of human
alteration of the biosphere is an increase in emerging infectious
diseases \cite{jones2008, fisher2012}. Such diseases pose a severe
threat to wildlife populations \cite{daszak2000}, and have caused
dramatic declines and extinctions in a wide range of taxa, including
echinoderms, mammals, birds, and amphibians
\cite{hewson2014, samuel2015, scheele2019, cunningham2021}. Amphibians
are experiencing particularly devastating impacts of disease due to the
recent emergence and global spread of the highly virulent amphibian
chytrid fungus, \emph{Batrachochytrium dendrobatidis} (Bd)
\cite{luedtke2023, scheele2019}. By one estimate, hundreds of species
have experienced Bd-caused declines, and numerous susceptible taxa are
extinct in the wild \cite{scheele2019}. These impacts to global
biodiversity may be unprecedented, and highlight the importance of
understanding mechanisms of species persistence in the presence of
emerging diseases \cite{russell2020}.

Evidence of natural recovery in the many Bd-impacted amphibian
populations is surprisingly rare \cite[for notable exceptions,
see][]{scheele2017, voyles2018, knapp2016}, suggesting that
disease-caused declines will be difficult to reverse. This apparent low
resilience to disease effects may be due to the limited ability of many
amphibians to develop Bd resistance and/or tolerance, which in turn,
could also lessen the effectiveness of potential Bd mitigation
strategies. Following pathogen arrival in a host population, resistance
(ability to limit pathogen burden) and tolerance (ability to limit the
harm caused by a particular burden) are key mechanisms to reduce disease
impacts \cite{raberg2009} and facilitate population persistence and
recovery \cite{brannelly2021}. Host immunity and evolution both play
important roles in the development of resistance and tolerance, and
utilizing these factors would seem a promising approach to developing
effective strategies to mitigate disease impacts in the wild
\cite{garner2016, woodhams2011}. However, several aspects of the
amphibian-Bd system present difficult obstacles, including (i) the
general inability of amphibians to mount an effective immune response
against Bd infection \cite{rosenblum2012, fites2013, grogan2018a}, and
(ii) the apparent rarity of evolution of more resistant/tolerant
genotypes \cite[but see][]{savage2016, grogan2018b}. These factors
suggest that reintroduction of amphibians into sites to reestablish
populations extirpated by Bd will often result not in population
recovery, but instead in the rapid reinfection and mortality of the
introduced animals and/or their progeny
\cite{hammond2021, knapp2022, stockwell2008, scheele2021}. If true, the
future of many amphibian species threatened by Bd appears bleak.

The mountain yellow-legged (MYL) frog, composed of the sister species
\emph{Rana muscosa} and \emph{Rana sierrae} \cite{vredenburg2007}, is
emblematic of the global declines of amphibians caused by Bd
\cite{scheele2019}. Once the most common amphibian in the high
elevation portion of California's Sierra Nevada mountains
\cite[USA,][]{grinnell1924}, during the past century this frog has
disappeared from more than 90\% of its historical range
\cite{vredenburg2007}. Due to the severity of its decline and the
increasing probability of extinction, both species are now listed as
``endangered'' under the U.S. Endangered Species Act. In the Sierra
Nevada, this decline was initiated by the introduction of non-native
trout into the extensive historically-fishless region
\cite{bradford1989, knapp2000} starting in the late 1800s. The arrival
of Bd in the mid-1900s and its subsequent spread \cite{vredenburg2019}
caused additional large-scale population extirpations
\cite{vredenburg2010, rachowicz2006}. These Bd-caused declines are
fundamentally different from the fish-caused declines because fish
eradication is feasible \cite{knapp1998} and results in the rapid
recovery of frog populations \cite{knapp2007, vredenburg2004}. In
contrast, Bd appears to persist in habitats even in the absence of
amphibian hosts \cite{walker2007}, and therefore represents a long-term
alteration of invaded ecosystems that amphibians will need to overcome
to reestablish populations.

Despite the catastrophic impact of Bd on MYL frogs, wherein most
Bd-naive populations are extirpated following Bd arrival
\cite{vredenburg2010}, some populations have persisted after epizootics
\cite[during which Bd infection intensity on frogs is very
high,][]{briggs2010} and are now recovering
(Figure~\ref{fig-recovery-model}) \cite{knapp2016}. Frogs in these
recovering populations show reduced susceptibility to Bd infection
\cite{knapp2016}, with infection intensity (``load'') on adults
consistently in the low-to-moderate range
\cite{briggs2010, knapp2011, joseph2018}. This reduced susceptibility
is evident even under controlled laboratory conditions
\cite{knapp2016}, indicative of host resistance against Bd infection
(and not simply an effect of factors external to individual frogs, e.g.,
environmental conditions). In addition to frogs from recovering
populations having higher resistance to Bd infection than those from
naive populations, they could also have higher tolerance, but no data
are currently available to evaluate this possibility. Therefore, we
focus on resistance throughout this paper. The observed resistance of
MYL frogs could be the result of several non-mutually exclusive
mechanisms, including natural selection for more resistant genotypes
\cite{savage2016, grogan2018b}, acquired immunity \cite{grogan2018a},
and/or inherent between-population differences that pre-date Bd
exposure. The possible evolution of MYL frog resistance and subsequent
population recovery is consistent with that expected under
``evolutionary rescue'', whereby rapid evolutionary change increases the
frequency of adaptive alleles and restores positive population growth
\cite{carlson2014, searle2020}. This intriguing possibility also
suggests an opportunity to expand recovery beyond the spatial scale
possible under natural recovery by utilizing resistant frogs from
recovering populations in reintroductions to vacant habitats
(Figure~\ref{fig-recovery-model}) \cite{joseph2018, mendelson2019}.

@@

Bd loads were fairly consistent before versus after translocation, and
loads were nearly always well below the level indicative of severe
chytridiomycosis (i.e., the disease caused by Bd) and associated frog
mortality (Figure S2)
\cite{joseph2018, vredenburg2010}. Although it is possible that the
observed relatively small changes in load are a consequence of
individuals with high Bd loads dying and therefore being unavailable for
sampling during the post-translocation period, the fact that there was
little difference in pre- versus post-translocation Bd loads even in
those populations that had very high frog survival (70556, 74976 - see
below; Figure S2) suggests a true lack of
substantial change in Bd load.

The ultimate measure of reintroduction success is the establishment of a
self-sustaining population. Given that it can take years or even decades
to determine the self-sustainability of a reintroduced population
\cite[for an example in MYL frogs, see][]{joseph2018}, the use of
proxies is essential for providing shorter-term insights into
reintroduction success and the factors driving it. Results from our CMR
surveys allowed us to accurately estimate frog survival, including over
the entire CMR time series for each site and during only the 1-year
period immediately following translocation. These estimates were made
using site-specific models analyzed using the mrmr package. We use these
estimates to describe general patterns of frog survival in all
translocated cohorts, and in an among-site meta-analysis of frog
survival to identify important predictors of 1-year frog survival (e.g.,
Bd load).

@@

Results from the frog translocation study indicated that most
populations showed evidence of successful reproduction and recruitment,
and that adult survival was often relatively high (described above).
Although suggestive of population establishment, a decade or more of
surveys may be necessary to confirm that populations are in fact
self-sustaining \cite{joseph2018}. To extend our inferences of
population establishment beyond those possible from the site-specific
CMR data, we developed a population viability model. Specifically, to
test whether the observed yearly adult survival probabilities in
translocated populations were sufficient for long-term viability, we
built a stage-structured matrix model that captured known frog
demography and included demographic and environmental stochasticity. We
parameterized the model using CMR data from translocated populations and
known life history values in this system (Table S1).

@@

To determine whether MYL frog populations show genomic patterns
consistent with an evolutionary response to Bd, we compared frog exomes
(i.e., coding region of a genome) between populations with contrasting
histories of Bd exposure. Specifically, we compared frog genomes sampled
in 4 populations that have not yet experienced a Bd-caused epizootic
(``naive'') \cite{zhou2015} versus in 5 populations that experienced a
Bd epizootic during the past several decades and have since recovered to
varying degrees (``recovering''; Figure~\ref{fig-allelefrequencies})
\cite{knapp2016, vredenburg2010}. Bd-exposure histories of the 9 study
populations are based on 10-20 years of VES and Bd surveillance using
skin swabbing \cite[e.g.,][]{knapp2016, zhou2015, wilber2022}. Naive
populations are characterized by large numbers of adults (i.e, typically
1000s), Bd prevalence that is generally 0\% except during occasional Bd
failed invasions \cite[during which Bd loads remain very
low,][]{wilber2022}, and no history of Bd epizootics since we first
surveyed these populations in the late 1990s and early 2000s
\cite{zhou2015}. In contrast, recovering populations exist in an
enzootic state \cite{briggs2010}, characterized by smaller numbers of
adults (generally \textless{} 500), high Bd prevalence \cite[often
\textgreater{} 80\%,][]{knapp2011}, and, in adults, moderate Bd loads
that are typically well below the level expected to cause mortality
\cite{vredenburg2010}. Naive and recovering populations can be
identified unambiguously using these differences in Bd prevalence and
load. Finally, there is no potential for frog dispersal between the 9
study populations due to intervening distances and topography, as well
as the presence of introduced (predatory) fish and fish-induced habitat
fragmentation. (See \textbf{Supporting Information - Frog evolution in
response to Bd - Study design} for additional details regarding the
study design.)

@@

Individual frogs clustered into 3 separate groups in principal component
space (Figure S6A), and clusters reflected
the species split (i.e., \emph{R. muscosa} versus \emph{R. sierrae}) and
the strong signature of isolation-by-distance that is characteristic of
MYL frogs \cite{rothstein2020, poorten2017, byrne2023}. Importantly,
each cluster contained at least one population from both the naive and
recovering groups, allowing us to distinguish allelic associations of
individuals sampled in the 2 population types versus allelic
associations resulting from population structure and genetic drift.

@@

The splined window analysis identified 33 outlier regions for
\(\pi_{diff}\) and 58 outlier regions for \emph{F\textsubscript{ST}}
(Figure~\ref{fig-spline-manhattan}A, B). Of these, 9 regions were
outliers for both metrics (``shared regions'') and 2 of these shared
regions also contained one or more of the outlier SNPs described above.
A total of 35 annotated genes were found in the 9 shared regions. Given
this large number of genes, here we focus on those with the strongest
signal of selection and/or immune-related functions. The largest
\(\pi_{diff}\), indicative of directional selection, occurred in a 163kb
region on Contig19, 12.9Mb upstream of the RIN3 outlier SNP
(Figure~\ref{fig-spline-manhattan}C). This region contains
approximately 500 SNPs and one annotated gene called
``interferon-induced very large GTPase 1-like'' (GVINP1). Additionally,
a shared outlier region on Contig1 contained two complement factor genes
(C6 and C7). Interestingly, this region had a large negative
\(\pi_{diff}\) value, consistent with balancing selection. Finally, one
shared outlier region on Contig8 contained one outlier SNP (TCF19) and
was within 360kb of another outlier SNP (VARS)
(Figure~\ref{fig-spline-manhattan}D, Figure S7).
This region (854kb from the beginning of the outlier window to the VARS
SNP) contained a total of 8 annotated genes. In \emph{Xenopus}, five of
these genes occur in the extended major histocompatibility complex (MHC)
Class I region (FLOT1, TUBB, MDC1, CCHCR1, TCF19) and three occur in the
extended MHC Class III region (HSP70, LSM2, VARS) \cite{ohta2006}.
Therefore, this region under selection is part of the extended MHC Class
I and III complex and shows synteny with other amphibian genomes.

@@

Disease-induced population declines are decimating global biodiversity
\cite{daszak2000}, but broadly-applicable strategies to recover
affected species are generally lacking \cite[e.g.,][]{garner2016}.
Here, we tested the possibility that populations of resistant
individuals from naturally recovering populations can be used to
reestablish extirpated populations of the endangered MYL frog in the
presence of a highly virulent fungal pathogen (Bd). Our results indicate
(i) the capacity of reintroduced populations to become established and
eventually recover despite ongoing disease, (ii) that the recovering
populations are likely to persist over a 50-year period, (iii) that
there are substantial genomic differences between naive and recovering
MYL frog populations, consistent with evolutionary change in frogs
following Bd exposure, and (iv) that some of the genomic regions under
selection contain genes related to disease resistance. Collectively,
these results (Figure~\ref{fig-recovery-model}) provide a rare example
of amphibian recovery in the presence of Bd, and have important
implications for the conservation and recovery of amphibians and other
taxa worldwide that are endangered by escalating impacts from emerging
infectious diseases. In light of the generally low success rate of
amphibian reintroduction efforts \cite{dodd2005}, our success in
reestablishing MYL frog populations via translocation of resistant
individuals is striking, and even more so given that MYL frogs were
driven to near-extinction by Bd.

In the following discussion, we follow the sequence of frog recovery
described in Figure~\ref{fig-recovery-model} to structure our key
points. Previous field studies in MYL frogs show that frog-Bd dynamics
and frog survival in the presence of Bd are fundamentally different
between naive and recovering populations. Following the arrival and
establishment of Bd in previously-naive populations, adult frogs develop
high Bd loads that lead to mass die-offs \cite{vredenburg2010}. In
contrast, in recovering populations adult frogs typically have
low-to-moderate and relatively constant Bd loads and mass die-offs are
not observed \cite[see also Figure S2]{briggs2010, knapp2011}. The differences in Bd load of frogs from
naive and recovering populations are also observed in controlled
laboratory studies \cite[see Figure 4 in][]{knapp2016}, and clearly
indicate that frogs from recovering populations exhibit resistance
against Bd infection. This resistance could in theory be due to several
factors, including natural selection for more resistant genotypes,
acquired immunity, and/or inherent between-population differences that
pre-date Bd exposure, but until now evidence to evaluate the role of
evolution was lacking.

@@

Importantly, some genomic regions that we identified as under selection
are associated with cellular and immunological mechanisms known to
contribute to disease resistance, including in amphibians
\cite{zamudio2020}. For example, the MHC plays an important role in
immunity. In our study, we identified a region that shows evidence of
selection in recovering populations and contains eight genes associated
with either the MHC Class I or Class III regions. These results
corroborate numerous previous studies linking MHC genes to amphibian
resistance against Bd \cite[e.g.,][]{savage2011, bataille2015}.
Similarly, the region with the strongest indication of directional
selection (as measured by \(\pi_{diff}\)) contains the
interferon-related gene GVINP1. Several previous studies of amphibians
have found this gene to be differentially expressed during Bd infection
\cite[e.g.,][]{grogan2018b, ellison2014} and in populations differing
in Bd susceptibility \cite{grogan2018b}. This gene is also strongly
linked to disease in salmon, explaining a notable 20\% of the resistance
phenotype \cite{robledo2020, robledo2018}. We also identified a region,
characterized by high \emph{F\textsubscript{ST}} and low \(\pi_{diff}\),
that contained the complement genes C6 and C7. The complement system
plays an important role in innate immunity \cite{rieraromo2016}, and
our results could indicate that balancing selection is acting in this
region of the genome to favor a diverse set of alleles, as is known for
C6 in humans \cite{soejima2005}. Based on the analysis of individual
outlier variants, the RIN3 gene showed a consistent pattern of allele
frequency differences across all nine of the frog populations sampled in
this study, indicating consistent selection in populations distributed
across a wide geographic area. This gene is associated with immune
response and in \emph{Xenopus} is expressed during appendage
regeneration \cite{fukazawa2009}. Finally, the outlier variant with the
lowest p-value was the uncharacterized gene LOC108802036. In the genome
of another frog species, this gene is located adjacent to a type I
interferon gene (Np-IFNi2) \cite{gan2018}, and together with GVINP1
further suggests the importance of interferon-related genes in this
system. Collectively, the genes associated with these genomic
differences may confer at least some degree of resistance against Bd
infection, an attribute that may be critically important to population
reestablishment and recovery in the presence of Bd.

@@

Results from the population viability model were also encouraging. In
particular, translocated populations with \textgreater{} 50\% survival
in the first year post-translocation were predicted to have a low
probability of extinction over 50 years (probability of extinction \(<\)
0.5 when year-1 juvenile survival probability was greater than 0.10).
The viability model highlighted the important role of frog survival in
affecting long-term population viability, and allowed us to extend the
temporal scale of our study beyond the years covered by our
post-translocation surveys. These long-term forecasts are important,
given that reintroduced MYL frog populations may often take decades to
achieve our ultimate goal of self-sustainability \cite{joseph2018}.
Making well-supported projections about the long-term outcome of
reintroduction efforts from shorter-term information is critically
important to the process of adaptive management of species
reintroduction programs \cite{seddon2007}, including the one we are
carrying out for MYL frogs. Specifically, the combined results from our
reintroduction study and viability modeling indicate that survival of
frogs in the first year following translocation is an effective proxy of
longer-term survival and population viability. In addition, given the
repeatability of frog survival at a site, 1-year frog survival also
serves as an effective proxy of site quality (i.e., the ability of a
site to support high frog survival and a viable frog population over the
long term). This proxy of site quality is important in the MYL frog
system because accurately predicting the ability of a site to support a
viable frog population \emph{a priori} remains difficult, even after
conducting 24 translocations over 16 years.
